---
title: "11_30_22_lab"
author: "Andy Tong"
format: html
---

Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
pertussis_cases_by_year <- data.frame(
                          Year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,
                                   2019L),
  No..Reported.Pertussis.Cases = c(107473,
                                   164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617)
)

```
```{r}
library(ggplot2)
ggplot(pertussis_cases_by_year) +
  aes(Year, No..Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  labs(x = "Year", y="Number of Cases", title = "Pertussis Cases by Year (1922 - 2019)")
```
Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(pertussis_cases_by_year) +
  aes(Year, No..Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  labs(x = "Year", y="Number of Cases", title = "Pertussis Cases by Year (1922 - 2019)") +
  geom_vline(xintercept = c(1946, 1996),  colour=c("blue", "red"), linetype = "longdash")+
  geom_text(aes(x = 1946+2, y = 250000, label = "wP"), colour = "blue")+
  geom_text(aes(x = 1996+2, y = 250000, label = "aP"), colour = "red")

```
Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine there was a slight uptick in cases of Pertussis. It did take a while though, this gives credence to maybe it was due to other factors such as testing becoming more sensitive and reporting more cases. Another possibility is the vaccine is slightly weaker but gives fewer side effects in a cost benefit analysis concluding that it is better. Another explanation is if the vaccine isn't updating often enough, the bacterial evolution could outpace vaccine development, creating new strains that could get around vaccination. A final explanation is that with time, people begin fearing the whooping cough less as they have never seen anyone get it. Because of this, they don't vaccinate their children for it as it doesn't seem like a problem. This could lead to a rise in cases. Between the fall in cases to almost 0 to the uptick, there are about 30 years, enough for one generation to grow up without fear or respect of the disease.

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
```
```{r}
head(subject, 3)
```
Q4. How many aP and wP infancy vaccinated subjects are in the dataset?
```{r}
table(subject$infancy_vac)
```
47 aP infancy vaccinated subjects and 49 wP infancy vaccinated subjects

Q5. How many Male and Female subjects/patients are in the dataset?
```{r}
table(subject$biological_sex)
```
66 females and 30 males are in the dataset

Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?
```{r}
table(subject$race, subject$biological_sex)
```
Here is the breakdown

```{r}
library(lubridate)
```
Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?
```{r}
subject$age <- today() - ymd(subject$year_of_birth)
```
```{r}
library(dplyr)

aP <- subject %>% filter(infancy_vac == "aP")
round(summary(time_length(aP$age, "years")))
```
```{r}
wP <- subject %>% filter(infancy_vac == "wP")
round(summary(time_length(wP$age, "years")))
```
Q8. Determine the age of all individuals at time of boost?
```{r}
time_diff_recieved_boost <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(time_diff_recieved_boost, "year")
head(age_at_boost)
```
Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?
```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
  fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) 
```
Yes they look significantly different. In the past people seem to get the vaccine whenever they could, but not at a young age. This leads to a distribution ranging from 25ish to almost 60. With the new vaccine everyone gets it before 30. 
```{r}
# Complete the API URLs...
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/ab_titer", simplifyVector = TRUE) 
```
Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- full_join(specimen, subject)
dim(meta)
head(meta)
```
Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```
Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```
Q12. What do you notice about the number of visit 8 specimens compared to other visits?
```{r}
table(abdata$visit)
```
There are very few visit 8 specimens compared to all the other visits.


```{r}
ig1 <- abdata %>% filter(isotype == "IgG1", visit!=8)
head(ig1)
```
Q13. Complete the following code to make a summary boxplot of Ab titer levels for all antigens:
```{r}
ggplot(ig1) +
  aes(MFI, antigen) +
  geom_boxplot() + 
  facet_wrap(vars(visit), nrow=2)+
  labs(x = "ab_titer") 
```
Q14. What antigens show differences in the level of IgG1 antibody titers recognizing them over time? Why these and not others?

FIM2/3 seems the most unique in terms of the level of IgG1 antibody titers. I didn't find anything on the website so I'm not entirely sure what that means. I am somewhat confused. 





